Path Analysis - Chapter 2

### 2016-01-09 sem_development.Rmd
Remove richness/diversity code to other script.
I also revised names of the rainfall and temperature data objects.

#### Load data
```{r pairs helper functions, echo=FALSE}
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}
```
```{r load ecological data, echo=FALSE}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")
# Vegetation data
veg_us_2005 <- read.csv("analysis/data/understory_veg_2005.csv")
veg_us_2011 <- read.csv("analysis/data/understory_veg_2011.csv")
# Untagged dbh data
untagged_dbh <- read.csv("analysis/data/untagged_dbh_summary.csv")
canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")
```
```{r load physical data, echo=FALSE, message=F}
# Physical environment: topography
library(rgdal); library(plyr); library(dplyr)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
plots_env <- plots_shp@data
summary(plots_env)
plots_env <- select(plots_env, -cat, -X, -Y)

@@

#### Join remotely sensed and derived physical data
LiDAR products, variables derived from DEMs
```{r join remotely sensed data}
plots_env <- inner_join(plots_env, plots_vegtopo, by = "PLOT_ID")
summary(plots_env)
plots_env$elev_m <- plots_env$elev15m * .3048
# plots_env <- select(plots_env, -elevation, -tmi) #remove elevation, tmi from old derivation (I think a legacy 10m USGS DEM)
plots_env$veght_m <- plots_env$veght15m * .3048
rm(plots_vegtopo)
```

@@

Examine temperature data, which is in long format and covers sample years 2004 through 2014. This means that I will need to exclude disease-related sampling from 2003.
```{r examine temperature & rain data}
summary(plots_temps)
summary(rain_days)
summary(rain_total)
unique(plots_temps$plotid)# 201 levels
unique(rain_days$plotid)# 202 levels
unique(rain_total$plotid)# 202 levels
tempsid <- as.character(unique(plots_temps$plotid))
rainid <- as.character(unique(rain_days$plotid))
rainid %in% tempsid
rainid[62]# BUSH01 still in the rain data
rain_days <- filter(rain_days, plotid != "BUSH01")
rain_total <- filter(rain_total, plotid != "BUSH01")
# Temperature and rainfall data now have same number of observations
rm(tempsid); rm(rainid)
```

@@

# BUSH01 plot included. I think we have limited ecological data for this plot, and no microclimate data. So, I am going to exclude it from all the data sets
# plots_env <- filter(plots_env, plotid != "BUSH01")
stems <- rename(stems, plotid = plot) %>% filter(plotid != "BUSH01")

plots_temps %>% filter(is.na(hrs_abv25))
# I am replacing the NAs with `0`, assuming true observations.
plots_temps[is.na(plots_temps)] <- 0
```

#### Save data with BUSH01 excluded
```{r save data}
write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
write.csv(rain_data, "analysis/data/rain_data_long.csv", row.names = F)
write.csv(plots_temps, "analysis/data/temp_data_long.csv", row.names = F)
write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
rm(plots_temps); rm(plots_rain)
```

#### Reload data that has BUSH01 excluded
```{r reload data}
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
rain_data <- read.csv("analysis/data/rain_data_long.csv")
temps_data <- read.csv("analysis/data/temp_data_long.csv")
stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

#### Plot contributing area vs. slope
This is a suggestion from Ryan to plot the contributing area vs. the slope for each of the plots in order to tease out how either influences infection and hosts. Decomposing TWI to determine which factor has a greater influence on disease dynamics. 
```{r decomposing TWI influence}
library(ggplot2)
# summary(plots_vegtopo)
# par(mfrow = c(3,1))
qplot(slope15m, flowacc15m, data = plots_env) +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m < 20), 
      main = "Flow Accumulation < 20") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 20 & flowacc15m < 1000), main = "Flow Accumulation 20 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 1000), 
      main = "Flow Accumulation > 1000") +
      geom_text(aes(label = plotid))

qplot(slope15m, facc_ws15m, data = plots_env)
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m < 15), 
      main = "facc < 15") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 15 & facc_ws15m < 50), 
      main = "facc 15 - 50") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 50 & facc_ws15m < 200), 
      main = "facc 50 - 200") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 200 & facc_ws15m < 1000), 
      main = "facc 200 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 1000), 
      main = "facc >1000") +
      geom_text(aes(label = plotid))

@@

#### Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
unique(plots_umca$plot)
umca_plots <- filter(plots_umca, plot != "BUSH01")
rm(plots_umca)
summary(bay_laurel) # This has stem-level data on bay laurel.
unique(bay_laurel$plot)
bay_laurel <- filter(bay_laurel, plot != "BUSH01")
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts}
umca_plots <- left_join(umca_plots, bay_laurel %>% filter(status == "Alive") %>% group_by(plot, year) %>%  summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(umca_plots)
```

```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

So now I have a plot-level of UMCA infection levels for each year. I want to join this with some of the environmental data.
```{r join env and umca data}
summary(plots_env)
plots_env$avg_psi <- rowMeans(cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
summary(plots_umca)
# plot == plotid, year == sample_year
plots_umca <- left_join(select(plots_umca, -species), plots_env, by = c("plot" = "plotid"))
summary(plots_umca)
filter(plots_umca, is.na(easting))
# NAs due to no topo data for abandoned SUGAR02
plots_umca <- na.omit(plots_umca)

plots_temps <- as.tbl(plots_temps)
plots_temps <- plots_temps %>% rename(year = sample_year)
plots_temps$plotid <- tolower(plots_temps$plotid)

plots_umca <- left_join(plots_umca, plots_temps, by = c("plotid", "year"))
```

#### Community vegetation data and variables
Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.

```{r calculate understory richness}
summary(veg_us_2005)
veg_us_2005 <- as.tbl(select(veg_us_2005, -X))
summary(veg_us_2011)
veg_us_2011 <- as.tbl(select(veg_us_2011, -X))
veg_us_2011
veg_us_2005

veg <- rbind(veg_us_2005, veg_us_2011)
unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)
unique(veg$us_species)

# I want to match the species to those used by Sarah in her second chapter
local_woody <- read.csv("analysis/data/presence_local_woody_n197.csv")
summary(local_woody)
unique(local_woody$spp.local)
woody_spp <- as.character(local_woody$spp.local)
woody_spp[woody_spp == "prunus"] <- "prunus sp."
sort(woody_spp) # only 29 species
sort(unique(veg$us_species))
# It looks like I will want to add a few more to her list

filter(veg, us_species == "ribes sp.")
filter(veg, us_species == "ribes")
veg$us_species[veg$us_species == "ribes"] <- "ribes sp."

woody_spp <- c(woody_spp, "ceanothus sp.", "cebe", "cecu", "frla", "garrya sp.", "gemo", "juca", "mafu", "mane", "mimulus sp", "pisa", "quag x quke", "qube", "quercus sp.", "quwi", "rhamnus sp.", "rhcr", "ribes sp.", "rica", "risa", "roca", "rosa sp.", "rubus sp.", "rudi", "salix sp.", "syal", "vaccinium sp.")
veg_sub <- filter(veg, veg$us_species %in% woody_spp)
veg_sub <- droplevels(veg_sub)

richness <- veg_sub %>%
      group_by(plotid, year) %>% 
      summarise(us_rich = length(unique(us_species)))
richness
```

Next, overstory richness from the DBH data for tagged & untagged stems.

```{r calculate overstory richness}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)
unique(untagged_dbh$plotid)
untagged_dbh <- filter(untagged_dbh, plotid != "bush01")
untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species)))

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)
stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
unique(stems$plotid)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)
stems <- droplevels(stems)

live_tagged <- stems %>% filter(status == "Alive") %>% group_by(plotid, year, species) %>% summarise(live_count = length(species))
unique(live_tagged$species)
os_richness <- rbind(live_tagged, select(untagged_dbh, plotid, species, year, live_count))

os_richness <- left_join(
      stems %>% filter(year == 2005 | year == 2014) %>% 
            group_by(plotid, year) %>% 
            summarise(tagged_rich = length(unique(species))),
      untagged_dbh %>% 
            group_by(plotid, year) %>% 
            summarise(untagged_rich = length(unique(species))), by = c("plotid", "year")) %>% 
      mutate(os_rich = rowSums(cbind(tagged_rich,untagged_rich), na.rm = T))
os_richness

richness <- left_join(os_richness, richness, by = c("plotid","year"))
richness
summary(richness)
```

>>>>

#### Load data
```{r pairs helper functions, echo=FALSE}
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}
```
```{r load ecological data, echo=FALSE}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")

# Vegetation transect data
veg_us_2005 <- read.csv("analysis/data/understory_veg_2005.csv")
veg_us_2011 <- read.csv("analysis/data/understory_veg_2011.csv")

# Untagged dbh data
untagged_dbh_2005 <- read.csv("analysis/data/untagged_dbh2005_summary.csv")
untagged_dbh_2014 <- read.csv("analysis/data/untagged_dbh2014_summary.csv")
canopy_cover <- read.csv("analysis/data/plot_canopy_cover_dens.csv")
```
```{r load physical data, echo=FALSE, message=F}
# Physical environment: topography
library(rgdal); library(plyr); library(dplyr)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
plots_env <- plots_shp@data
summary(plots_env)
plots_env <- select(plots_env, -cat, -X, -Y)

@@

#### Join remotely sensed and derived physical data
LiDAR products, variables derived from DEMs
```{r join remotely sensed data}
plots_env <- inner_join(plots_env, plots_vegtopo, by = "PLOT_ID")
summary(plots_env)
plots_env$elev_m <- plots_env$elev15m * .3048
# plots_env <- select(plots_env, -elevation, -tmi) #remove elevation, tmi from old derivation (I think a legacy 10m USGS DEM)
plots_env$veght_m <- plots_env$veght15m * .3048
rm(plots_vegtopo)

```

@@

Examine temperature data, which is in long format and covers sample years 2004 through 2014. This means that I will need to exclude disease-related sampling from 2003.
```{r examine temperature & rain data}
summary(rs_temps)
summary(summer_temps)

summary(ppt_days)
summary(ppt_total)
rain_df <- merge(ppt_days@data, ppt_total@data)
summary(rain_df)

unique(rs_temps$plotid)# 201 levels
unique(summer_temps$plotid)# 201 levels
unique(rain_df$plotid)# 202 levels

tempsid <- as.character(unique(rs_temps$plotid))
rainid <- as.character(unique(rain_df$plotid))
rainid %in% tempsid
rainid[62]# BUSH01 still in the rain data
rain_df <- filter(rain_df, plotid != "BUSH01")

# Temperature and rainfall data now have same number of observations
rm(tempsid); rm(rainid)
```

@@

# BUSH01 plot included. I think we have limited ecological data for this plot, and no microclimate data. So, I am going to exclude it from all the data sets
# plots_env <- filter(plots_env, plotid != "BUSH01")
stems <- as.tbl(stems)
stems
stems <- select(stems, -X) 
stems <- rename(stems, plotid = plot) %>% filter(plotid != "BUSH01")

rs_temps %>% filter(is.na(hrs_abv25))
# I am replacing the NAs with 0, assuming true observations, i.e. there were zero hours where temperatures were above 25C
rs_temps[is.na(rs_temps)] <- 0
```

#### Save data with BUSH01 excluded
```{r save data}
write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
write.csv(rain_df, "analysis/data/rain_data_long.csv", row.names = F)
write.csv(rs_temps, "analysis/data/rs_temps_data.csv", row.names = F)
write.csv(summer_temps, "analysis/data/summer_temps_data.csv", row.names = F)
write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
rm(plots_temps); rm(plots_rain)
```

### Reload data that has BUSH01 excluded
```{r reload data}
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
rain_df <- read.csv("analysis/data/rain_data_long.csv")
rs_temps <- read.csv("analysis/data/rs_temps_data.csv")
summer_temps <- read.csv("analysis/data/summer_temps_data.csv")
stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

### Plot contributing area vs. slope
This is a suggestion from Ryan to plot the contributing area vs. the slope for each of the plots in order to tease out how either influences infection and hosts. Decomposing TWI to determine which factor has a greater influence on disease dynamics. 
```{r decomposing TWI influence}
library(ggplot2)
# summary(plots_vegtopo)
# par(mfrow = c(3,1))
names(plots_env)
qplot(flowacc15m, facc_ws15m, data = plots_env) #+
      geom_text(aes(label = plotid))

qplot(slope15m, flowacc15m, data = plots_env) #+
      geom_text(aes(label = plotid))

qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m < 20), 
      main = "Flow Accumulation < 20") #+
      geom_text(aes(label = plotid))

qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 20 & flowacc15m < 1000), main = "Flow Accumulation 20 - 1000") #+
      geom_text(aes(label = plotid))

qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 1000), 
      main = "Flow Accumulation > 1000") #+
      geom_text(aes(label = plotid))

# I think the 'facc_ws15m' is the estimate from the 'r.watershed' function in GRASS GIS
qplot(slope15m, facc_ws15m, data = plots_env)
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m < 15), 
      main = "facc < 15") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 15 & facc_ws15m < 50), 
      main = "facc 15 - 50") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 50 & facc_ws15m < 200), 
      main = "facc 50 - 200") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 200 & facc_ws15m < 1000), 
      main = "facc 200 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 1000), 
      main = "facc >1000") +
      geom_text(aes(label = plotid))

@@

#### Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.

unique(plots_umca$plot)
umca_plots <- filter(plots_umca, plot != "BUSH01")
rm(plots_umca)
summary(bay_laurel) # This has stem-level data on bay laurel.
unique(bay_laurel$plot)
bay_laurel <- filter(bay_laurel, plot != "BUSH01")
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year.
```{r calculate plot level leaf counts}
umca_plots <- left_join(
      umca_plots, bay_laurel %>% filter(status == "Alive") %>% 
            group_by(plot, year) %>% 
            summarise(avg_lfct = mean(slc, na.rm = T),
                      tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(umca_plots)
```

Explore the NAs in the UMCA leaf counts.
```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `umca_plots` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

So now I have a plot-level of UMCA infection levels for each year in terms of symptomatic leaf counts and number of un/infected stems. I want to join this with some of the plot-level environmental data.
```{r join env and umca data}
summary(plots_env)

# Calculate a mean potential solar radiation
plots_env$avg_psi <- rowMeans(
      cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
summary(umca_plots)
# plot == plotid, year == sample_year

# Join the environmental data to the bay laurel infection data
umca_plots <- left_join(select(umca_plots, -species), plots_env, by = c("plot" = "plotid"))
summary(umca_plots)
filter(umca_plots, is.na(lon_utm))
# NAs due to no topo data for abandoned SUGAR02; remove SUGAR02
umca_plots <- na.omit(umca_plots)
umca_plots <- rename(umca_plots, plotid = plot, sample_year = year)
umca_plots$plotid <- tolower(umca_plots$plotid)

# Join rainy season and summer temperatures to bay laurel infection data
rs_temps <- as.tbl(rs_temps)
rs_temps
rs_temps$plotid <- tolower(rs_temps$plotid)

summer_temps <- as.tbl(summer_temps)
summer_temps
summer_temps$plotid <- tolower(summer_temps$plotid)

umca_plots <- left_join(umca_plots, rs_temps, by = c("plotid", "sample_year"))
umca_plots <- left_join(umca_plots, summer_temps, by = c("plotid", "sample_year"))
umca_plots
```
